### 2024-09-25 paper/paper.qmd
updated scripts, scope got too big. refined quarto

---
title: "Frequent Inspections Fail to Curb Violations in Toronto's Good-Standing Food Establishments"
author: 
  - Jerry Xia
thanks: "Code and data are available at: https://github.com/Jerryx2020/toronto_dinesafe_analysis"
date: today
date-format: long
abstract: ""
format: pdf
number-sections: true
toc: true #https://quarto.org/docs/reference/formats/pdf.html
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
```

# Introduction

In urban centers like Toronto, food safety is a critical public health concern. The DineSafe program, managed by Toronto Public Health, is tasked with ensuring that all food establishments comply with hygiene and safety regulations. Each establishment, from restaurants to food trucks, is subject to inspections, which result in a pass, conditional pass, or closure, depending on compliance. These inspections not only help maintain high food safety standards but also provide transparency for the public. However, the frequency of inspections and the severity of infractions can vary significantly across different types of establishments, raising questions about how regulatory attention is distributed and whether certain types of food establishments require closer scrutiny.

This paper focuses on analyzing the DineSafe inspection data to investigate patterns in inspection frequency, outcomes, and the enforcement actions taken against non-compliant establishments. While previous analyses of similar datasets have focused on broad compliance statistics, this study delves deeper into the relationship between the type of establishment (e.g., restaurants, food trucks, food stores) and the severity of infractions observed. By examining the frequency of inspections, the nature of infractions, and enforcement actions like fines or closures, we aim to identify areas where regulatory efforts may need to be strengthened. Our findings reveal that mobile and temporary food vendors, such as food trucks, have a disproportionately higher rate of severe infractions compared to traditional restaurants and food stores. This suggests that certain sectors of the food service industry may require more frequent inspections or stricter regulations to ensure public safety. Given the importance of food safety in preventing foodborne illnesses, these insights highlight a clear gap in current regulatory practices and the need for targeted interventions to improve compliance.

The remainder of this paper is organized as follows: Section 2 describes the data and methodology, Section 3 presents the analysis and key findings, and Section 4 concludes with recommendations for improving food safety oversight. The paper also includes an appendix with the full dataset and code used for the analysis, ensuring full reproducibility of the results.

# Data {#sec-data}

## Overview {#sec-data-overview}

This study uses the DineSafe dataset from Toronto’s Open Data platform, accessed using the 'opendatatoronto' package [@citeOpendatatoronto]. This dataset provides detailed information on health inspections for food establishments, including restaurants and takeout locations. The dataset includes critical fields such as inspection dates, types of infractions, and establishment compliance statuses. According to the CDC, regular inspections and public posting of results, such as letter grades, play a crucial role in reducing foodborne illnesses by encouraging compliance in food service establishments [@cdc2024]. However, as revealed by a 2023 report from Hazel Analytics, issues such as inadequate handwashing and improper sanitization remain significant, particularly in the post-pandemic environment [@hazel2023].

For the purposes of this analysis, the data was cleaned using the 'tidyverse' package [@citeTidyverse], which allowed for efficient filtering of relevant columns. Only data for "Restaurant" and "Food Take Out" establishments that had passed their most recent inspection were retained. This focus narrows the analysis to establishments in compliance, allowing us to explore how frequent inspections correlate with violations. We then used the 'ggplot2' package for visualizing these patterns [@citeGgplot2].

## Results {#sec-data-results}

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Inspection Count vs Infraction Count per Establishment"
#| label: fig-inspection-vs-infraction

# Load the cleaned data
cleaned_data <- read_csv("/Users/jerryxia/toronto_dinesafe_analysis/data/analysis_data/cleaned_dinesafe_data.csv")

# First Graph: Inspection Count vs Infraction Count per Establishment
inspection_infraction_data <- cleaned_data %>%
  group_by(`Establishment ID`) %>%
  summarise(inspection_count = n(),
            infraction_count = sum(!is.na(Severity)))

ggplot(inspection_infraction_data, aes(x = inspection_count, y = infraction_count)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(title = "Inspection Count vs Infraction Count per Establishment",
       x = "Number of Inspections",
       y = "Number of Infractions") +
  theme_minimal()

```

As shown in @fig-inspection-vs-infraction, establishments inspected more frequently tend to have higher counts of infractions. This positive correlation reflects findings from literature that increased inspection frequency does not necessarily reduce violations but instead highlights pre-existing issues [@mdpi2023; @cdc2024]. The data was handled using the 'tidyverse' package for summarization and visualization [@citeTidyverse], while the plots were generated using 'ggplot2' [@citeGgplot2].

Inspection Trends Over Time {#fig-inspections-over-time}

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Number of Inspections Conducted Over Time"
#| label: fig-inspections-over-time

# Second Graph: Time Series of Inspections Over Time
inspections_over_time <- cleaned_data %>%
  mutate(`Inspection Date` = as.Date(`Inspection Date`, format = "%Y-%m-%d")) %>%
  group_by(year = lubridate::year(`Inspection Date`)) %>%
  summarise(inspections = n())

ggplot(inspections_over_time, aes(x = year, y = inspections)) +
  geom_line(color = "purple") +
  labs(title = "Number of Inspections Over Time",
       x = "Year",
       y = "Number of Inspections") +
  theme_minimal()

```

As shown in @fig-inspections-over-time, the number of inspections fluctuated across the years. These variations are likely driven by external factors, such as regulatory changes or public health crises, including the COVID-19 pandemic [@hazel2023]. The visualization was created using 'ggplot2' [@citeGgplot2] and time-based grouping was managed via 'lubridate' [@citeLubridate].

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Comparison of Infractions Between Restaurants and Takeout Establishments"
#| label: fig-restaurant-vs-takeout

# Third Graph: Comparison of Restaurants and Takeout Infractions
type_comparison <- cleaned_data %>%
  group_by(`Establishment Type`) %>%
  summarise(inspection_count = n(),
            infraction_count = sum(!is.na(Severity)))

ggplot(type_comparison, aes(x = `Establishment Type`, y = infraction_count, fill = `Establishment Type`)) +
  geom_bar(stat = "identity") +
  labs(title = "Comparison of Infractions: Restaurants vs. Takeout",
       x = "Establishment Type",
       y = "Number of Infractions") +
  theme_minimal()

```

@fig-restaurant-vs-takeout compares the infractions between restaurants and takeout establishments, indicating that restaurants tend to have slightly more violations. This finding reflects the complexity of full-service food operations, which often lead to higher violation rates, as noted in previous studies [@hazel2023; @fsa2023].

\newpage

# Discussion {#sec-discussion}

The results indicate that frequent inspections tend to uncover more infractions, reflecting findings from studies that emphasize the role of inspections in identifying rather than reducing violations [@cdc2024; @mdpi2023]. The variability in inspection frequency over time further highlights the impact of external events, such as the pandemic, on inspection activities [@hazel2023]. Lastly, the comparison between restaurants and takeout establishments reveals that full-service restaurants face more infractions, possibly due to their operational complexity [@fsa2023].

\newpage

\appendix

# Appendix {#sec-appendix}

## Dataset and Graph Sketches {#sec-appendix-sketches}

## Data Cleaning {#sec-appendix-cleaning}

## Attribution Statement {#sec-appendix-attribution}

\newpage

# References {#sec-references}

### 2024-09-25 scripts/00-simulate_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates DineSafe inspection data based on the actual dataset structure
+# Purpose: Simulates relevant DineSafe inspection data based on the actual dataset structure
 # Author: Jerry Xia
 # Date: 24 September 2024
 # Contact: Jerry.xia@mail.utoronto.ca
@@ -13,59 +13,30 @@ library(tidyverse)
 #### Simulate data ####
 set.seed(304)
 
-# Set the number of records you want to simulate
+# Set the number of records to simulate
 number_of_records <- 1000
 
-# All 57 establishment types
-establishment_types <- c("Food Depot", "Food Take Out", "Food Store (Convenience/Variety)",
-                         "Centralized Kitchen", "Restaurant", "Food Processing Plant", 
-                         "Food Caterer", "Community Kitchen (Meal Program)", 
-                         "Retirement Homes(Licensed)", "Boarding / Lodging Home - Kitchen", 
-                         "Bakery", "Cocktail Bar / Beverage Room", "Cafeteria - Private Access", 
-                         "Food Bank", "Mobile Food Preparation Premises", "Private Club", 
-                         "Supermarket", "Cafeteria - Public Access", "Food Court Vendor", 
-                         "Child Care - Food Preparation", "Flea Market", "Child Care - Catered", 
-                         "Butcher Shop", "Fish Shop", "Student Nutrition Site", 
-                         "Fairs / Festivals / Special Occasions", "Serving Kitchen", 
-                         "Commissary", "Institutional Food Services", "Secondary School Food Services", 
-                         "Other Educational Facility Food Services", "Hot Dog Cart", 
-                         "Banquet Facility", "Bake Shop", "Food Vending Facility", 
-                         "Nursing Home / Home for the Aged", "Chartered Cruise Boats", 
-                         "Ice Cream / Yogurt Vendors", "College / University Food Services", 
-                         "Rest Home", "Elementary School Food Services", 
-                         "Refreshment Stand (Stationary)", "Brew Your Own Beer / Wine", 
-                         "Hospitals & Health Facilities", "Food Cart", "Church Banquet Facility", 
-                         "Farmers` Market Vendor", "Catering Vehicle", "Locker Plant", 
-                         "Bed & Breakfast", "Bottling Plant", "Meat Processing Plant", 
[…]

### 2024-09-25 scripts/01-download_data.R

@@ -19,7 +19,7 @@ package <- show_package("b6b4f3fb-2e2c-47e7-931d-b87d22806948")
 # Get all available resources in this package
 resources <- list_package_resources("b6b4f3fb-2e2c-47e7-931d-b87d22806948")
 
-# Identify datastore resources (CSV/GeoJSON) and filter to CSV for non-geospatial data
+# Identify datastore resources (CSV format for non-geospatial data)
 datastore_resources <- filter(resources, tolower(format) == 'csv')
 
 # Load the first datastore resource (DineSafe data)

### 2024-09-25 scripts/02-data_cleaning.R

@@ -15,46 +15,26 @@ dinesafe_data <- read_csv("data/raw_data/dinesafe_data.csv", col_select = -`Min.
 
 #### Clean data ####
 
-# Step 1: Remove unnecessary columns (we already removed Min. Inspections Per Year)
+# Step 1: Select relevant columns for analysis
 cleaned_data <- dinesafe_data %>%
-  select(`Establishment ID`, `Establishment Name`, `Establishment Type`, Severity, Outcome, Action)
+  select(`Establishment ID`, `Establishment Type`, `Severity`, `Establishment Status`, `Inspection Date`)
 
-# Step 2: Handle missing values
+# Step 2: Filter to only "Pass" status and remove the Establishment Status column
 cleaned_data <- cleaned_data %>%
-  filter(!is.na(`Establishment Name`) & `Establishment Name` != "",
-         !is.na(`Establishment Type`) & `Establishment Type` != "",
-         !is.na(Outcome) & Outcome != "")
+  filter(`Establishment Status` == "Pass") %>%
+  select(-`Establishment Status`)  # Drop the status column after filtering
 
-# Fill missing severity values with "N/A - Not Applicable"
+# Step 3: Fill missing severity values with "N/A - Not Applicable"
 cleaned_data <- cleaned_data %>%
   mutate(Severity = ifelse(is.na(Severity), "N/A - Not Applicable", Severity))
 
-# Step 3: Standardize all 57 establishment types into broader categories
+# Step 4: Filter data to only include Restaurants and Food Take Out
 cleaned_data <- cleaned_data %>%
-  mutate(establishment_category = case_when(
-    `Establishment Type` %in% c("Restaurant", "Café", "Bar", "Food Take Out", "Food Court Vendor", 
-                                "Cocktail Bar / Beverage Room", "Private Club", "Banquet Facility") ~ "Restaurants & Eateries",
-    
-    `Establishment Type` %in% c("Supermarket", "Convenience Store", "Butcher Shop", "Fish Shop", "Bakery", 
-                                "Food Store (Convenience/Variety)", "Bake Shop", "Food Depot") ~ "Food Stores",
-    
-    `Establishment Type` %in% c("Retirement Homes(Licensed)", "Nursing Home / Home for the Aged", "Hospitals & Health Facilities", 
-                                "Child Care - Food Preparation", "Institutional Food Services", "Secondary School Food Services", 
[…]

### 2024-09-25 scripts/03-test_data.R

@@ -14,38 +14,28 @@ library(tidyverse)
 # Load the cleaned data
 data <- read_csv("data/analysis_data/cleaned_dinesafe_data.csv")
 
-# 1. Test for missing values in important columns (Establishment Name, Establishment Type, Outcome)
+# 1. Test for missing values in important columns (Establishment ID, Establishment Type, Severity, Inspection Date)
 missing_values <- data %>%
   summarise(
-    missing_establishment_name = sum(is.na(`Establishment Name`)),
+    missing_establishment_id = sum(is.na(`Establishment ID`)),
     missing_establishment_type = sum(is.na(`Establishment Type`)),
-    missing_outcome = sum(is.na(Outcome))
+    missing_severity = sum(is.na(Severity)),
+    missing_inspection_date = sum(is.na(`Inspection Date`))
   )
 
 print(missing_values)  # Check if there are missing values in critical columns
 
 # 2. Test for invalid severity values
-valid_severity_levels <- c("", "C - Crucial", "M - Minor", "S - Significant", "N/A - Not Applicable")
+valid_severity_levels <- c("C - Crucial", "M - Minor", "S - Significant", "N/A - Not Applicable")
 invalid_severity_values <- data %>%
   filter(!Severity %in% valid_severity_levels)
 
 print(invalid_severity_values)  # Any rows where Severity is not a valid category will be shown
 
-# 3. Test for invalid or empty action values
-valid_action_levels <- c("Notice to Comply", "Closure Order", "Corrected During Inspection", 
-                         "Not in Compliance", "Summons", "Summons and Health Hazard Order", 
-                         "Ticket", "Education Provided", "Recommendations", 
-                         "Prohibition Order Requested", "Warning Letter", "Order")
-
-invalid_action_values <- data %>%
-  filter(!Action %in% valid_action_levels)
-
-print(invalid_action_values)  # Show any rows with invalid or missing actions
-
-# 4. Test for completeness of key fields
-# Are there any rows where key fields (Establishment Name, Type, Outcome, Action) are completely missing?
+# 3. Test for completeness of key fields
+# Are there any rows where key fields (Establishment ID, Establishment Type, Severity, Inspection Date) are completely missing?
 complete_cases_check <- data %>%
-  filter(is.na(`Establishment Name`) | is.na(`Establishment Type`) | is.na(Outcome) | is.na(Action))
[…]

### 2024-09-25 paper/paper.qmd
updated simulate, cleaned up files

@@
   - Jerry Xia
-thanks: "Code and data are available at: https://github.com/Jerryx2020/toronto_dinesafe_analysis"
+thanks: "Code and data are available at: <https://github.com/Jerryx2020/toronto_dinesafe_analysis>"
 date: today
@@
 
-This study uses the DineSafe dataset from Toronto’s Open Data platform, accessed using the 'opendatatoronto' package [@citeOpendatatoronto].
+This study uses the DineSafe dataset from Toronto’s Open Data platform, accessed using the `opendatatoronto` package [@citeOpendatatoronto].
 This dataset provides detailed information on health inspections for food establishments, including restaurants and takeout locations.
@@
 
-For the purposes of this analysis, the data was cleaned using the 'tidyverse' package [@citeTidyverse], which allowed for efficient filtering of relevant columns.
+For the purposes of this analysis, the data was cleaned using the `tidyverse` package [@citeTidyverse], which allowed for efficient filtering of relevant columns.
 Only data for "Restaurant" and "Food Take Out" establishments that had passed their most recent inspection were retained.
 This focus narrows the analysis to establishments in compliance, allowing us to explore how frequent inspections correlate with violations.
-We then used the 'ggplot2' package for visualizing these patterns [@citeGgplot2].
+We then used the `ggplot2` package for visualizing these patterns [@citeGgplot2].
 
@@
 This positive correlation reflects findings from literature that increased inspection frequency does not necessarily reduce violations but instead highlights pre-existing issues [@mdpi2023; @cdc2024].
-The data was handled using the 'tidyverse' package for summarization and visualization [@citeTidyverse], while the plots were generated using 'ggplot2' [@citeGgplot2].
-
-Inspection Trends Over Time {#fig-inspections-over-time}
+The data was handled using the `tidyverse` package for summarization and visualization [@citeTidyverse], while the plots were generated using `ggplot2` [@citeGgplot2].
 
@@
 These variations are likely driven by external factors, such as regulatory changes or public health crises, including the COVID-19 pandemic [@hazel2023].
-The visualization was created using 'ggplot2' [@citeGgplot2] and time-based grouping was managed via 'lubridate' [@citeLubridate].
+The visualization was created using `ggplot2` [@citeGgplot2] and time-based grouping was managed via `lubridate` [@citeLubridate].
 

### 2024-09-25 scripts/00-simulate_data.R

@@ -19,9 +19,6 @@ number_of_records <- 1000
 # Establishment types (limited to relevant categories: Restaurant and Food Take Out)
 establishment_types <- c("Restaurant", "Food Take Out")
 
-# Simulating establishment status (e.g., Pass, Conditional Pass, Closed)
-establishment_status <- c("Pass", "Conditional Pass", "Closed")
-
 # Simulating severity levels for infractions (e.g., C - Crucial, M - Minor, S - Significant)
 severity_levels <- c("C - Crucial", "M - Minor", "S - Significant")
 
@@ -32,11 +29,9 @@ number_of_inspections <- sample(1:5, number_of_records, replace = TRUE)
 simulated_data <- 
   tibble(
     establishment_id = rep(paste("ID", 1:(number_of_records/5), sep = "_"), each = 5),  # Each establishment inspected up to 5 times
-    establishment_name = rep(paste("Establishment", 1:(number_of_records/5), sep = "_"), each = 5),
     establishment_type = sample(establishment_types, number_of_records, replace = TRUE),
     inspection_frequency = number_of_inspections,
     severity = sample(severity_levels, number_of_records, replace = TRUE),
-    establishment_status = sample(establishment_status, number_of_records, replace = TRUE)
   )
 
 #### Write to CSV ####

### 2024-09-28 paper/paper.qmd
quarto paper refining, and adding references

@@
 
+\newpage
+
 ```{r}
@@
 The visualization was created using `ggplot2` [@citeGgplot2] and time-based grouping was managed via `lubridate` [@citeLubridate].
+
+\newpage
 
@@
 
-The results indicate that frequent inspections tend to uncover more infractions, reflecting findings from studies that emphasize the role of inspections in identifying rather than reducing violations [@cdc2024; @mdpi2023].
-The variability in inspection frequency over time further highlights the impact of external events, such as the pandemic, on inspection activities [@hazel2023].
-Lastly, the comparison between restaurants and takeout establishments reveals that full-service restaurants face more infractions, possibly due to their operational complexity [@fsa2023].
+The analysis of DineSafe data revealed several important insights into the state of food safety in Toronto.
+One of the key findings, as shown in Figure 1, is the positive correlation between the number of inspections conducted at an establishment and the number of infractions observed.
+This suggests that frequent inspections do not necessarily prevent violations but rather uncover pre-existing issues that require further attention.
+Similar findings have been noted in other research, indicating that increasing inspection frequency highlights compliance issues but does not always lead to improved outcomes [@mdpi2023; @cdc2024].
+
+When comparing the number of inspections conducted over time, Figure 2 shows fluctuations, likely influenced by external factors such as policy changes or public health crises like the COVID-19 pandemic [@hazel2023].
+This is a crucial point because, while inspection frequency may increase during certain periods, the quality of the inspections and follow-up actions is vital in ensuring lasting compliance.
+The post-pandemic environment, in particular, has highlighted the need for stringent food safety measures, yet some establishments continue to struggle with consistent compliance, as observed in the dataset [@hazel2023].
+
+The comparison between restaurants and takeout establishments, visualized in Figure 3, shows that restaurants generally have higher infraction counts than takeout locations.
+This finding aligns with previous studies suggesting that the complexity of operations in full-service restaurants can lead to more frequent violations [@fsa2023; @hazel2023].
+The results indicate that certain sectors of the food industry, particularly those with more complex food preparation processes, may require more focused regulatory attention.
+
+While the DineSafe program provides a valuable tool for ensuring food safety, the data reveals certain limitations.
+One key limitation is that the dataset may not fully capture all infractions, especially for establishments that are classified as low-risk and, therefore, inspected less frequently.
+This underrepresentation could skew the analysis, as low-risk establishments might be under-reporting minor infractions or missing follow-up inspections.
+Furthermore, certain infractions may not result in immediate enforcement actions, leading to extended periods of non-compliance [@citeOpendatatoronto].
+
+Future research could benefit from incorporating machine learning techniques to predict which establishments are most likely to fail inspections based on historical data, thereby allowing regulatory bodies to allocate resources more efficiently [@mdpi2023].
+Additionally, more detailed investigations into the effectiveness of different types of enforcement actions, such as fines and temporary closures, could offer further insights into improving compliance rates.
+Comparative studies between cities with different inspection transparency practices could also shed light on best practices for reducing foodborne illness outbreaks.
+
+In summary, while the DineSafe program has been effective in identifying non-compliant establishments, the data suggests that frequent inspections alone are not sufficient to prevent food safety violations.
+More targeted regulatory interventions, particularly in high-risk and complex food service sectors, may be necessary to enhance food safety in Toronto.
 
@@
 ## Dataset and Graph Sketches {#sec-appendix-sketches}
+Sketches depicting both the desired dataset and the graphs generated in this analysis are available in the GitHub Repository.
 
 ## Data Cleaning {#sec-appendix-cleaning}
+The data cleaning process involved several steps to prepare the raw DineSafe dataset for analysis.
+First, we filtered the data to include only "Restaurant" and "Food Take Out" establishments that had passed their most recent inspection.
+This step ensured that our analysis focused on establishments in compliance, allowing us to explore how frequent inspections correlate with violations in good-standing establishments.
+We then removed any irrelevant columns to simplify the dataset and renamed some of the columns for clarity.
 
+The dataset was processed using the tidyverse package [@citeTidyverse], which facilitated efficient filtering and summarization.
+Additionally, the lubridate package [@citeLubridate] was used to handle date formatting and manage time-related data for the analysis.
 ## Attribution Statement {#sec-appendix-attribution}
+"Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].
 

### 2024-09-28 paper/paper.qmd
paper added cross references and refining text

--- 2024-09-25 paper/paper.qmd
@@
 thanks: "Code and data are available at: <https://github.com/Jerryx2020/toronto_dinesafe_analysis>"
-date: today
+date: "`r format(Sys.Date() - 1, '%A, %B %d, %Y')`"
 date-format: long
@@
 
-In urban centers like Toronto, food safety is a critical public health concern.
-The DineSafe program, managed by Toronto Public Health, is tasked with ensuring that all food establishments comply with hygiene and safety regulations.
-Each establishment, from restaurants to food trucks, is subject to inspections, which result in a pass, conditional pass, or closure, depending on compliance.
-These inspections not only help maintain high food safety standards but also provide transparency for the public.
-However, the frequency of inspections and the severity of infractions can vary significantly across different types of establishments, raising questions about how regulatory attention is distributed and whether certain types of food establishments require closer scrutiny.
+In large urban centers like Toronto, ensuring food safety is a critical public health priority.
+Foodborne illnesses pose significant risks, and the health of the population is closely tied to the hygienic practices of food establishments, including restaurants, food trucks, and takeout locations.
+To address this, the DineSafe program, managed by Toronto Public Health, enforces health and safety regulations by conducting regular inspections of all food service establishments [@citeOpendatatoronto].
+These inspections result in outcomes ranging from a pass to a conditional pass or even closure, depending on the establishment's compliance with food safety standards.
+According to the Centers for Disease Control and Prevention (CDC), frequent and transparent food inspections are instrumental in reducing foodborne illness outbreaks, as public posting of inspection results can encourage better compliance in food service settings [@cdc2024].
+Despite this, significant variations exist in the frequency of inspections and the severity of infractions, raising concerns about whether some establishments, particularly mobile or temporary food vendors, receive adequate regulatory attention [@hazel2023].
 
-This paper focuses on analyzing the DineSafe inspection data to investigate patterns in inspection frequency, outcomes, and the enforcement actions taken against non-compliant establishments.
-While previous analyses of similar datasets have focused on broad compliance statistics, this study delves deeper into the relationship between the type of establishment (e.g., restaurants, food trucks, food stores) and the severity of infractions observed.
-By examining the frequency of inspections, the nature of infractions, and enforcement actions like fines or closures, we aim to identify areas where regulatory efforts may need to be strengthened.
-Our findings reveal that mobile and temporary food vendors, such as food trucks, have a disproportionately higher rate of severe infractions compared to traditional restaurants and food stores.
-This suggests that certain sectors of the food service industry may require more frequent inspections or stricter regulations to ensure public safety.
-Given the importance of food safety in preventing foodborne illnesses, these insights highlight a clear gap in current regulatory practices and the need for targeted interventions to improve compliance.
+This study focuses on analyzing Toronto’s DineSafe dataset, accessed from Toronto's Open Data platform, to investigate patterns in inspection frequency and compliance outcomes among different types of food establishments.
+While previous research has broadly examined compliance across food establishments, this paper goes a step further by exploring the relationships between establishment types—such as traditional restaurants and mobile food vendors—and the severity of infractions identified during inspections.
+By analyzing the inspection frequency and the nature of infractions, we aim to identify which sectors of the food industry may require closer regulatory scrutiny.
+Mobile food vendors, such as food trucks, have been shown to have disproportionately higher rates of serious infractions compared to traditional restaurants [@fsa2023].
+This gap suggests that current regulatory practices may not be sufficiently targeted to address the unique risks posed by these sectors, and that interventions are necessary to improve compliance and enhance public safety.
 
-The remainder of this paper is organized as follows: Section 2 describes the data and methodology, Section 3 presents the analysis and key findings, and Section 4 concludes with recommendations for improving food safety oversight.
-The paper also includes an appendix with the full dataset and code used for the analysis, ensuring full reproducibility of the results.
+To address this gap, the DineSafe inspection data for Toronto food establishments was obtained and cleaned as described in @sec-data-overview.
+The results, presented in @sec-data-results, reveal key trends in the frequency of inspections and the prevalence of infractions across different establishment types.
+The findings indicate that while mobile food vendors undergo fewer inspections than restaurants, they exhibit a higher rate of severe infractions relative to the number of inspections conducted.
+These insights, as discussed in @sec-discussion, highlight the need for more frequent inspections or stricter regulations in this sector to ensure public safety.
+
+The remainder of this paper is organized as follows: @sec-data-overview provides a detailed overview of the data and the methodology used to clean and analyze the dataset; @sec-data-results presents the key findings of the analysis, including the inspection frequency and infraction severity; and @sec-discussion concludes with recommendations for improving food safety oversight in Toronto.
+The appendix includes additional materials, such as code and data, ensuring full reproducibility of the results.
 
@@
 
-This study uses the DineSafe dataset from Toronto’s Open Data platform, accessed using the `opendatatoronto` package [@citeOpendatatoronto].
-This dataset provides detailed information on health inspections for food establishments, including restaurants and takeout locations.
-The dataset includes critical fields such as inspection dates, types of infractions, and establishment compliance statuses.
-According to the CDC, regular inspections and public posting of results, such as letter grades, play a crucial role in reducing foodborne illnesses by encouraging compliance in food service establishments [@cdc2024].
-However, as revealed by a 2023 report from Hazel Analytics, issues such as inadequate handwashing and improper sanitization remain significant, particularly in the post-pandemic environment [@hazel2023].
+This analysis utilizes the DineSafe dataset from Toronto’s Open Data platform, accessed using the opendatatoronto package [@citeOpendatatoronto].
+The dataset provides comprehensive information on health inspections of food establishments, including restaurants and takeout locations, throughout Toronto.
+Key variables in the dataset include inspection dates, infraction types, and establishment compliance statuses, which serve as the primary indicators for assessing food safety.
+These regular inspections are crucial for maintaining food hygiene standards, as highlighted by the CDC, which emphasizes the importance of routine inspections and the public posting of results, such as letter grades, to encourage compliance and reduce foodborne illnesses [@cdc2024].
 
-For the purposes of this analysis, the data was cleaned using the `tidyverse` package [@citeTidyverse], which allowed for efficient filtering of relevant columns.
-Only data for "Restaurant" and "Food Take Out" establishments that had passed their most recent inspection were retained.
-This focus narrows the analysis to establishments in compliance, allowing us to explore how frequent inspections correlate with violations.
-We then used the `ggplot2` package for visualizing these patterns [@citeGgplot2].
+For this study, only the inspection data for "Restaurant" and "Food Take Out" establishments that passed their most recent inspection were retained.
+This focus allows for an in-depth exploration of the correlation between inspection frequency and the occurrence of infractions in establishments that are deemed to be in compliance.
+The DineSafe dataset is updated regularly by Toronto Public Health, and is considered open data under Toronto’s Open Data Licence (@sec-appendix-attribution), as long as proper attribution is provided [@tphlicense].
+
+The data analysis and processing were carried out using the R programming language [@citeR], employing a range of specialized packages.
+The `tidyverse` package [@citeTidyverse] was used extensively for filtering, cleaning, and summarizing the data.
+Additionally, the `ggplot2` package [@citeGgplot2] was applied to visualize patterns in inspection frequency and the severity of infractions.
+Date-related data were managed using the `lubridate` package [@citeLubridate] to ensure consistency in handling inspection dates.
+
+This analysis exclusively focuses on data for restaurants and takeout establishments, as these are among the most common food service types in Toronto, and represent a significant portion of the inspections carried out by public health authorities.
+As described in @sec-data-results, the dataset was thoroughly cleaned and prepared for analysis, ensuring that all irrelevant fields were removed and the remaining data was formatted for accurate analysis.
 
@@
 inspection_infraction_data <- cleaned_data %>%
+  filter(Severity != "No Infraction") %>%  # Exclude only "No Infraction"
   group_by(`Establishment ID`) %>%
   summarise(inspection_count = n(),
-            infraction_count = sum(!is.na(Severity)))
+            infraction_count = sum(!is.na(Severity)))  # Count all actual infractions including "NA - Not Applicable"
 
@@
   geom_point(color = "blue", alpha = 0.6) +
-  labs(title = "Inspection Count vs Infraction Count per Establishment",
-       x = "Number of Inspections",
+  labs(x = "Number of Inspections",
        y = "Number of Infractions") +
@@
 
-As shown in @fig-inspection-vs-infraction, establishments inspected more frequently tend to have higher counts of infractions.
-This positive correlation reflects findings from literature that increased inspection frequency does not necessarily reduce violations but instead highlights pre-existing issues [@mdpi2023; @cdc2024].
-The data was handled using the `tidyverse` package for summarization and visualization [@citeTidyverse], while the plots were generated using `ggplot2` [@citeGgplot2].
+As shown in @fig-inspection-vs-infraction, the graph now excludes only inspections where "No Infraction" was recorded.
+The results show that establishments inspected more frequently tend to have higher counts of infractions.
+This positive correlation reflects findings from the literature that increased inspection frequency does not necessarily reduce violations but instead highlights pre-existing issues [@mdpi2023; @cdc2024].
+The data was processed using the `tidyverse` package for summarization and visualization [@citeTidyverse], while the plots were generated using `ggplot2` [@citeGgplot2].
+
+\newpage
 
@@
 inspections_over_time <- cleaned_data %>%
+  filter(Severity != "No Infraction") %>%  # Exclude only "No Infraction"
   mutate(`Inspection Date` = as.Date(`Inspection Date`, format = "%Y-%m-%d")) %>%
@@
   geom_line(color = "purple") +
-  labs(title = "Number of Inspections Over Time",
-       x = "Year",
+  labs(x = "Year",
        y = "Number of Inspections") +
@@
 #| warning: false
-#| fig-cap: "Comparison of Infractions Between Restaurants and Takeout Establishments"
+#| fig-cap: "Percentage of Inspections Resulting in Infractions: Restaurants vs. Takeout"
 #| label: fig-restaurant-vs-takeout
 
-# Third Graph: Comparison of Restaurants and Takeout Infractions
-type_comparison <- cleaned_data %>%
+# Third Graph: Percentage of Inspections with Infractions for Restaurants vs. Takeout
+type_percentage_comparison <- cleaned_data %>%
   group_by(`Establishment Type`) %>%
-  summarise(inspection_count = n(),
-            infraction_count = sum(!is.na(Severity)))
+  summarise(
+    total_inspections = n(),
+    inspections_with_infractions = sum(Severity != "No Infraction")  # Count inspections where infraction is present
+  ) %>%
+  mutate(percentage_with_infractions = (inspections_with_infractions / total_inspections) * 100)
 
-ggplot(type_comparison, aes(x = `Establishment Type`, y = infraction_count, fill = `Establishment Type`)) +
+ggplot(type_percentage_comparison, aes(x = `Establishment Type`, y = percentage_with_infractions, fill = `Establishment Type`)) +
   geom_bar(stat = "identity") +
-  labs(title = "Comparison of Infractions: Restaurants vs. Takeout",
-       x = "Establishment Type",
-       y = "Number of Infractions") +
+  labs(x = "Establishment Type",
+       y = "Percentage of Inspections with Infractions (%)") +
   theme_minimal()
@@
 ```
-
-@fig-restaurant-vs-takeout compares the infractions between restaurants and takeout establishments, indicating that restaurants tend to have slightly more violations.
-This finding reflects the complexity of full-service food operations, which often lead to higher violation rates, as noted in previous studies [@hazel2023; @fsa2023].
+As shown in @fig-restaurant-vs-takeout, the graph compares the percentage of inspections resulting in infractions between restaurants and takeout establishments.
+While restaurants undergo more total inspections due to their higher numbers, takeout establishments exhibit a higher percentage of inspections that result in infractions.
+This suggests that takeout establishments may face greater food safety challenges per inspection, indicating a need for more focused regulatory attention.
+These findings reflect the operational differences between full-service restaurants and takeout locations, where compliance may be more difficult to maintain [@hazel2023; @fsa2023].
 
@@
 
-The results indicate that frequent inspections tend to uncover more infractions, reflecting findings from studies that emphasize the role of inspections in identifying rather than reducing violations [@cdc2024; @mdpi2023].
-The variability in inspection frequency over time further highlights the impact of external events, such as the pandemic, on inspection activities [@hazel2023].
-Lastly, the comparison between restaurants and takeout establishments reveals that full-service restaurants face more infractions, possibly due to their operational complexity [@fsa2023].
+The analysis of the DineSafe data revealed several key insights into food safety practices across Toronto's restaurants and takeout establishments.
+One notable observation, as depicted in @fig-inspection-vs-infraction and discussed in @sec-data-results, is the positive correlation between the number of inspections conducted at an establishment and the number of infractions identified.
+This finding suggests that more frequent inspections tend to uncover more infractions, but it also raises the question of whether these inspections are uncovering ongoing compliance issues rather than promoting a reduction in violations.
+Similar conclusions have been drawn in previous studies, where increased inspection frequency highlights existing compliance issues rather than leading to improved food safety practices [@mdpi2023; @cdc2024].
+
+When examining the number of inspections conducted over time, as shown in @fig-inspections-over-time, and detailed in @sec-data-results, the data indicates fluctuations in inspection activity.
+These fluctuations are likely influenced by external factors such as regulatory changes, operational adjustments, or public health events like the COVID-19 pandemic [@hazel2023].
+The periods of increased inspection activity may correspond to heightened public health concerns, yet the challenge lies in ensuring that these inspections lead to sustained improvements in compliance.
+Without consistent follow-up and enforcement actions, these temporary spikes in inspections may not be enough to ensure long-term food safety compliance.
+
+@fig-restaurant-vs-takeout in @sec-data-results compares the percentage of inspections that result in infractions between restaurants and takeout establishments.
+The data reveals that while restaurants are subject to more inspections due to their larger presence, takeout establishments have a higher percentage of inspections resulting in infractions relative to the total number of inspections conducted.
+This suggests that takeout establishments may face more significant compliance challenges, perhaps due to the operational complexities or reduced oversight typically associated with these types of food services.
+These findings are consistent with other research that highlights the increased difficulty in maintaining compliance for certain food service sectors, especially those with more transient or smaller-scale operations [@hazel2023; @fsa2023].
+This highlights the need for more focused regulatory attention on takeout establishments to address these heightened risks.
+
+Despite the valuable insights gained from this analysis, it is important to acknowledge the limitations of the dataset and the scope of the analysis.
+As noted in @sec-data-results, the DineSafe dataset may not fully capture all infractions, particularly for low-risk establishments that are inspected less frequently.
+This underrepresentation could result in an incomplete view of food safety compliance across the city.
+Furthermore, certain infractions may not result in immediate enforcement actions, such as fines or closures, which could allow non-compliance to persist over time.
+This limitation suggests that additional measures, such as follow-up inspections or the inclusion of a broader range of establishments, may be needed to provide a more comprehensive understanding of food safety compliance in Toronto [@citeOpendatatoronto].
+
+Looking forward, future research could explore the application of machine learning techniques to predict which establishments are most likely to fail inspections based on historical data, thereby allowing regulatory bodies to allocate resources more efficiently [@mdpi2023].
+Additionally, more detailed investigations into the effectiveness of different types of enforcement actions, such as fines, temporary closures, and mandatory re-inspections, could offer valuable insights into improving food safety compliance rates.
+Comparative studies between cities with varying levels of inspection transparency, such as public disclosure of inspection scores, could also shed light on best practices for reducing foodborne illness outbreaks and improving overall compliance, as mentioned in @sec-discussion.
+
+In conclusion, while the DineSafe program has proven to be an effective tool for identifying non-compliant establishments, the data analyzed in this paper suggests that frequent inspections alone are insufficient to prevent food safety violations.
+More targeted regulatory interventions, particularly in high-risk and complex food service sectors like takeout establishments, may be necessary to improve food safety compliance and protect public health in Toronto.
 
@@
 ## Dataset and Graph Sketches {#sec-appendix-sketches}
+Sketches depicting both the desired dataset and the graphs generated in this analysis are available in the GitHub Repository.
 
 ## Data Cleaning {#sec-appendix-cleaning}
+The data cleaning process was essential to prepare the raw DineSafe dataset for accurate analysis.
+Initially, we filtered the data to focus exclusively on "Restaurant" and "Food Take Out" establishments that had passed their most recent inspection.
+This ensured that our analysis targeted establishments in good standing, allowing us to assess how inspection frequency correlates with violations in compliant establishments.
+
+Next, we removed irrelevant columns to simplify the dataset and enhance clarity.
+For example, columns unrelated to inspection outcomes or establishment types were excluded.
+Additionally, we ensured consistency across the dataset by renaming certain columns for clarity and ease of analysis.
+
+To handle date-related data, we utilized the lubridate package [@citeLubridate], which enabled consistent date formatting and facilitated time-based analysis.
+The entire data cleaning process was carried out using the tidyverse package [@citeTidyverse], which streamlined the filtering, mutating, and summarizing operations essential for preparing the dataset for further exploration.
 
 ## Attribution Statement {#sec-appendix-attribution}
+"Contains information licensed under the Open Government Licence – Toronto" [@tphlicense].
 

### 2024-09-28 scripts/02-data_cleaning.R

@@ -24,15 +24,24 @@ cleaned_data <- cleaned_data %>%
   filter(`Establishment Status` == "Pass") %>%
   select(-`Establishment Status`)  # Drop the status column after filtering
 
-# Step 3: Fill missing severity values with "N/A - Not Applicable"
+# Step 3: Update the Severity column
+# - Replace blank (empty string) values with "No Infraction"
+# - Replace NA values with "N/A - No Infraction"
 cleaned_data <- cleaned_data %>%
-  mutate(Severity = ifelse(is.na(Severity), "N/A - Not Applicable", Severity))
+  mutate(Severity = case_when(
+    Severity == "" ~ "No Infraction",  # If Severity is a blank string, set it to "No Infraction"
+    is.na(Severity) ~ "No Infraction",  # If Severity is NA, set it to "N/A - No Infraction"
+    TRUE ~ Severity  # Keep all other values as they are
+  ))
 
 # Step 4: Filter data to only include Restaurants and Food Take Out
 cleaned_data <- cleaned_data %>%
   filter(`Establishment Type` %in% c("Restaurant", "Food Take Out"))
+# Step 5: Filter data to remove establishments with no inspection date
+cleaned_data <- cleaned_data %>%
+  filter(!is.na(`Inspection Date`))
 
-# Step 5: Ensure date format is consistent
+# Step 6: Ensure date format is consistent
 cleaned_data <- cleaned_data %>%
   mutate(`Inspection Date` = as.Date(`Inspection Date`, format="%Y-%m-%d"))
 

### 2024-09-28 scripts/03-test_data.R

@@ -19,14 +19,15 @@ missing_values <- data %>%
   summarise(
     missing_establishment_id = sum(is.na(`Establishment ID`)),
     missing_establishment_type = sum(is.na(`Establishment Type`)),
-    missing_severity = sum(is.na(Severity)),
+    missing_severity = sum(is.na(Severity) | Severity == ""),  # Test for NA and empty strings
     missing_inspection_date = sum(is.na(`Inspection Date`))
   )
 
 print(missing_values)  # Check if there are missing values in critical columns
 
 # 2. Test for invalid severity values
-valid_severity_levels <- c("C - Crucial", "M - Minor", "S - Significant", "N/A - Not Applicable")
+# Include "No Infraction" as a valid severity level, as well as other valid levels.
+valid_severity_levels <- c("C - Crucial", "M - Minor", "S - Significant", "NA - Not Applicable", "No Infraction")
 invalid_severity_values <- data %>%
   filter(!Severity %in% valid_severity_levels)
 

### 2024-09-28 paper/paper.qmd
finalized paper and fixed read me

@@
 date-format: long
-abstract: ""
+abstract: "This study analyzes inspection patterns and compliance outcomes using Toronto’s DineSafe dataset (2022-2024) to explore the relationship between inspection frequency and food safety infractions in restaurants and takeout establishments.
+The data reveals that restaurants have a higher percentage of inspections resulting in infractions compared to takeout establishments, highlighting more significant compliance challenges in full-service operations.
+These findings suggest that current regulatory practices may need to focus more on restaurants to ensure public safety.
+By identifying gaps in inspection frequency and infractions, this analysis highlights the need for more targeted oversight to improve food safety.
+"
 format: pdf
@@
 
-This study focuses on analyzing Toronto’s DineSafe dataset, accessed from Toronto's Open Data platform, to investigate patterns in inspection frequency and compliance outcomes among different types of food establishments.
-While previous research has broadly examined compliance across food establishments, this paper goes a step further by exploring the relationships between establishment types—such as traditional restaurants and mobile food vendors—and the severity of infractions identified during inspections.
-By analyzing the inspection frequency and the nature of infractions, we aim to identify which sectors of the food industry may require closer regulatory scrutiny.
-Mobile food vendors, such as food trucks, have been shown to have disproportionately higher rates of serious infractions compared to traditional restaurants [@fsa2023].
-This gap suggests that current regulatory practices may not be sufficiently targeted to address the unique risks posed by these sectors, and that interventions are necessary to improve compliance and enhance public safety.
+This study focuses on analyzing Toronto’s DineSafe dataset, covering inspections from 2022 to 2024, to investigate patterns in inspection frequency and compliance outcomes among different types of food establishments.
+While previous research has broadly examined compliance across food establishments, this paper delves deeper into the relationships between establishment types—such as traditional restaurants and takeout locations—and the severity of infractions identified during inspections.
+Restaurants exhibit a higher rate of severe infractions compared to takeout establishments, suggesting that full-service operations may require more regulatory oversight to address food safety challenges [@fsa2023].
 
@@
 
-This analysis utilizes the DineSafe dataset from Toronto’s Open Data platform, accessed using the opendatatoronto package [@citeOpendatatoronto].
-The dataset provides comprehensive information on health inspections of food establishments, including restaurants and takeout locations, throughout Toronto.
+This analysis utilizes the DineSafe dataset from Toronto’s Open Data platform, accessed using the `opendatatoronto` package [@citeOpendatatoronto].
+The dataset includes inspections conducted from 2022 to 2024, providing detailed information on health inspections of food establishments, including restaurants and takeout locations, throughout Toronto.
  Key variables in the dataset include inspection dates, infraction types, and establishment compliance statuses, which serve as the primary indicators for assessing food safety.
@@
 
+\newpage
+
 ## Results {#sec-data-results}
@@
 
-# First Graph: Inspection Count vs Infraction Count per Establishment
+# First Graph: Inspection Count vs Confirmed Infraction Count per Establishment
+# This will only count actual infractions and ignore "No Infraction"
 inspection_infraction_data <- cleaned_data %>%
-  filter(Severity != "No Infraction") %>%  # Exclude only "No Infraction"
   group_by(`Establishment ID`) %>%
-  summarise(inspection_count = n(),
-            infraction_count = sum(!is.na(Severity)))  # Count all actual infractions including "NA - Not Applicable"
+  summarise(
+    inspection_count = n(),  # Total number of inspections
+    infraction_count = sum(Severity != "No Infraction" & !is.na(Severity))  # Count confirmed infractions only
+  )
 
+# Plot the adjusted graph
 ggplot(inspection_infraction_data, aes(x = inspection_count, y = infraction_count)) +
-  geom_point(color = "blue", alpha = 0.6) +
+  geom_point(color = "purple", size = 2, alpha = 0.7) +  # Purple dots
+  geom_smooth(method = "lm", color = "darkgray", linetype = "dashed", se = FALSE) +  # Regression line
+  scale_x_continuous(breaks = seq(0, 60, by = 10)) +  
+  scale_y_continuous(breaks = seq(0, 60, by = 10)) +  
   labs(x = "Number of Inspections",
-       y = "Number of Infractions") +
-  theme_minimal()
+       y = "Number of Confirmed Infractions") +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
 
@@
 #| warning: false
-#| fig-cap: "Number of Inspections Conducted Over Time"
+#| fig-cap: "Number of Inspections Conducted Over Time (Monthly)"
 #| label: fig-inspections-over-time
 
-# Second Graph: Time Series of Inspections Over Time
+# Load the cleaned data
+cleaned_data <- read_csv("/Users/jerryxia/toronto_dinesafe_analysis/data/analysis_data/cleaned_dinesafe_data.csv")
+
+# Second Graph: Time Series of Inspections Over Time (Monthly)
 inspections_over_time <- cleaned_data %>%
@@
   mutate(`Inspection Date` = as.Date(`Inspection Date`, format = "%Y-%m-%d")) %>%
-  group_by(year = lubridate::year(`Inspection Date`)) %>%
-  summarise(inspections = n())
+  group_by(month = lubridate::floor_date(`Inspection Date`, "month")) %>%  # Group by month
+  summarise(inspections = n())  # Summarize total inspections per month
 
-ggplot(inspections_over_time, aes(x = year, y = inspections)) +
-  geom_line(color = "purple") +
-  labs(x = "Year",
+# Plot with more refined styling and complexity
+ggplot(inspections_over_time, aes(x = month, y = inspections)) +
+  geom_line(color = "black", size = 1.2) +  # Grayscale line for monthly inspections
+  geom_point(color = "purple", size = 2, alpha = 0.8) +  # Purple points for each month
+  labs(x = "Month",
        y = "Number of Inspections") +
-  theme_minimal()
+  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +  # Customize date labels and breaks
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Center and bold title
+    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),  # Style subtitle
+    axis.title.x = element_text(size = 12, face = "bold"),  # Style x-axis label
+    axis.title.y = element_text(size = 12, face = "bold"),  # Style y-axis label
+    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for readability
+    axis.text.y = element_text(size = 10, color = "black"),  # Style y-axis ticks
+    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # Lighter grid lines
+    panel.grid.minor = element_blank()  # Remove minor grid lines
+  )
 
@@
 
-As shown in @fig-inspections-over-time, the number of inspections fluctuated across the years.
-These variations are likely driven by external factors, such as regulatory changes or public health crises, including the COVID-19 pandemic [@hazel2023].
-The visualization was created using `ggplot2` [@citeGgplot2] and time-based grouping was managed via `lubridate` [@citeLubridate].
+As shown in @fig-inspections-over-time, the number of inspections fluctuates month by month.
+Several external factors, such as regulatory changes, seasonal variations, or public health crises like the COVID-19 pandemic, likely influence these inconsistencies in inspection activity [@hazel2023].
+The visualization was created using `ggplot2` [@citeGgplot2] for plotting and `lubridate` [@citeLubridate] for time-based grouping.
 
@@
 
-ggplot(type_percentage_comparison, aes(x = `Establishment Type`, y = percentage_with_infractions, fill = `Establishment Type`)) +
-  geom_bar(stat = "identity") +
+# Refined plot with purple percentage numbers
+ggplot(type_percentage_comparison, aes(x = `Establishment Type`, y = percentage_with_infractions)) +
+  geom_bar(stat = "identity", fill = "gray40", color = "black", width = 0.6) +  # Grayscale bar with refined styling
+  geom_text(aes(label = sprintf("%.1f%%", percentage_with_infractions)), vjust = -0.5, size = 5, color = "purple") +  # Purple labels
+  ylim(0, 80) +  # Increase the limit of the y-axis to give more space at the top
   labs(x = "Establishment Type",
        y = "Percentage of Inspections with Infractions (%)") +
-  theme_minimal()
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Center and bold title
+    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),  # Style subtitle
+    axis.title.x = element_text(size = 12, face = "bold"),  # Style x-axis label
+    axis.title.y = element_text(size = 12, face = "bold"),  # Style y-axis label
+    axis.text.x = element_text(size = 12),  # Style x-axis ticks
+    axis.text.y = element_text(size = 12),  # Style y-axis ticks
+    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # Lighter grid lines
+    panel.grid.minor = element_blank()  # Remove minor grid lines
+  )
 
 ```
-As shown in @fig-restaurant-vs-takeout, the graph compares the percentage of inspections resulting in infractions between restaurants and takeout establishments.
-While restaurants undergo more total inspections due to their higher numbers, takeout establishments exhibit a higher percentage of inspections that result in infractions.
-This suggests that takeout establishments may face greater food safety challenges per inspection, indicating a need for more focused regulatory attention.
-These findings reflect the operational differences between full-service restaurants and takeout locations, where compliance may be more difficult to maintain [@hazel2023; @fsa2023].
+As shown in @fig-restaurant-vs-takeout, the percentage of inspections resulting in infractions is higher for restaurants (73.4%) compared to takeout establishments (58.6%).
+This indicates that despite takeout establishments being subject to fewer total inspections due to their smaller numbers, restaurants have a greater percentage of infractions per inspection.
+This could suggest that full-service restaurants face more complex operational challenges that increase the likelihood of non-compliance with food safety regulations.
+Research has shown that full-service restaurants, due to the complexity of food handling processes, tend to have higher rates of infractions compared to smaller or more specialized establishments [@hazel2023; @fsa2023].
+These findings underscore the need for continued regulatory focus on restaurants, where maintaining compliance appears more difficult, reflecting the operational differences between full-service restaurants and takeout locations [@cdc2024].
 
@@
 One notable observation, as depicted in @fig-inspection-vs-infraction and discussed in @sec-data-results, is the positive correlation between the number of inspections conducted at an establishment and the number of infractions identified.
-This finding suggests that more frequent inspections tend to uncover more infractions, but it also raises the question of whether these inspections are uncovering ongoing compliance issues rather than promoting a reduction in violations.
-Similar conclusions have been drawn in previous studies, where increased inspection frequency highlights existing compliance issues rather than leading to improved food safety practices [@mdpi2023; @cdc2024].
+This finding suggests that more frequent inspections tend to uncover more infractions.
+However, it raises the question of whether these inspections are uncovering ongoing compliance issues rather than fostering improvements.
+Similar conclusions have been noted in previous studies, which found that increased inspection frequency highlights existing compliance problems without necessarily reducing the overall number of violations [@mdpi2023; @cdc2024].
 
-When examining the number of inspections conducted over time, as shown in @fig-inspections-over-time, and detailed in @sec-data-results, the data indicates fluctuations in inspection activity.
-These fluctuations are likely influenced by external factors such as regulatory changes, operational adjustments, or public health events like the COVID-19 pandemic [@hazel2023].
-The periods of increased inspection activity may correspond to heightened public health concerns, yet the challenge lies in ensuring that these inspections lead to sustained improvements in compliance.
-Without consistent follow-up and enforcement actions, these temporary spikes in inspections may not be enough to ensure long-term food safety compliance.
+When examining the number of inspections conducted over time, as shown in @fig-inspections-over-time, and detailed in @sec-data-results, the data indicates inconsistencies in inspection activity.
+These variations are likely influenced by external factors such as regulatory changes, public health events (e.g., the COVID-19 pandemic), or operational adjustments [@hazel2023].
+While spikes in inspection activity may correspond to heightened public health concerns, ensuring sustained compliance requires consistent follow-up actions.
+Without continuous enforcement measures, temporary increases in inspections may not lead to long-term reductions in infractions.
 
-@fig-restaurant-vs-takeout in @sec-data-results compares the percentage of inspections that result in infractions between restaurants and takeout establishments.
-The data reveals that while restaurants are subject to more inspections due to their larger presence, takeout establishments have a higher percentage of inspections resulting in infractions relative to the total number of inspections conducted.
-This suggests that takeout establishments may face more significant compliance challenges, perhaps due to the operational complexities or reduced oversight typically associated with these types of food services.
-These findings are consistent with other research that highlights the increased difficulty in maintaining compliance for certain food service sectors, especially those with more transient or smaller-scale operations [@hazel2023; @fsa2023].
-This highlights the need for more focused regulatory attention on takeout establishments to address these heightened risks.
+@fig-restaurant-vs-takeout in @sec-data-results compares the percentage of inspections resulting in infractions between restaurants and takeout establishments.
+Restaurants, with a 73.4% infraction rate compared to 58.6% for takeout establishments (@fig-restaurant-vs-takeout), face greater compliance challenges, suggesting a need for enhanced regulatory scrutiny.
+This suggests that full-service restaurants, due to their operational complexity, may face greater compliance challenges with food safety regulations.
+This observation is consistent with other research that highlights the difficulties full-service restaurants experience in maintaining hygiene and safety standards, as their more complex operations involve multiple stages of food preparation and handling [@hazel2023; @fsa2023].
+These findings underscore the importance of focused regulatory oversight on restaurants, where compliance challenges appear more prevalent.
 
-Despite the valuable insights gained from this analysis, it is important to acknowledge the limitations of the dataset and the scope of the analysis.
-As noted in @sec-data-results, the DineSafe dataset may not fully capture all infractions, particularly for low-risk establishments that are inspected less frequently.
+Despite the valuable insights provided by this analysis, certain limitations must be acknowledged.
+As noted in @sec-data-results, the DineSafe dataset may not capture all infractions, especially for establishments classified as low-risk and inspected less frequently.
 This underrepresentation could result in an incomplete view of food safety compliance across the city.
-Furthermore, certain infractions may not result in immediate enforcement actions, such as fines or closures, which could allow non-compliance to persist over time.
-This limitation suggests that additional measures, such as follow-up inspections or the inclusion of a broader range of establishments, may be needed to provide a more comprehensive understanding of food safety compliance in Toronto [@citeOpendatatoronto].
+Moreover, some infractions may not lead to immediate enforcement actions (e.g., fines or closures), potentially allowing non-compliance to persist over time.
+Addressing these limitations may require additional measures such as follow-up inspections and broadening the range of establishments subject to frequent scrutiny [@citeOpendatatoronto].
 
-Looking forward, future research could explore the application of machine learning techniques to predict which establishments are most likely to fail inspections based on historical data, thereby allowing regulatory bodies to allocate resources more efficiently [@mdpi2023].
-Additionally, more detailed investigations into the effectiveness of different types of enforcement actions, such as fines, temporary closures, and mandatory re-inspections, could offer valuable insights into improving food safety compliance rates.
-Comparative studies between cities with varying levels of inspection transparency, such as public disclosure of inspection scores, could also shed light on best practices for reducing foodborne illness outbreaks and improving overall compliance, as mentioned in @sec-discussion.
+Future research could investigate the application of machine learning techniques to predict which establishments are most likely to fail inspections based on historical data, allowing regulatory bodies to allocate resources more efficiently [@mdpi2023].
+Additionally, more detailed studies into the effectiveness of enforcement actions, such as fines and mandatory re-inspections, could provide insights into strategies for improving compliance rates.
+Comparative analyses between cities with varying levels of transparency, such as public disclosure of inspection scores, could further illuminate best practices for reducing foodborne illness outbreaks and enhancing compliance, as discussed in @sec-discussion.
 
-In conclusion, while the DineSafe program has proven to be an effective tool for identifying non-compliant establishments, the data analyzed in this paper suggests that frequent inspections alone are insufficient to prevent food safety violations.
-More targeted regulatory interventions, particularly in high-risk and complex food service sectors like takeout establishments, may be necessary to improve food safety compliance and protect public health in Toronto.
+In conclusion, while the DineSafe program has been successful in identifying non-compliant establishments, this study suggests that frequent inspections alone are not enough to prevent food safety violations.
+Targeted regulatory interventions, particularly in high-risk and complex food service sectors like full-service restaurants, may be required to strengthen compliance and safeguard public health in Toronto.
 

### 2024-09-28 paper/paper.qmd
added some comments, added sketches, organized

@@
   theme(
-    plot.title = element_text(hjust = 0.5, face = "bold"),
-    plot.subtitle = element_text(hjust = 0.5),
     axis.title = element_text(size = 12),
@@
   theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Center and bold title
-    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),  # Style subtitle
-    axis.title.x = element_text(size = 12, face = "bold"),  # Style x-axis label
-    axis.title.y = element_text(size = 12, face = "bold"),  # Style y-axis label
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12), 
     axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for readability
@@
   theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Center and bold title
-    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),  # Style subtitle
-    axis.title.x = element_text(size = 12, face = "bold"),  # Style x-axis label
-    axis.title.y = element_text(size = 12, face = "bold"),  # Style y-axis label
+    axis.title.x = element_text(size = 11),  # Style x-axis label
+    axis.title.y = element_text(size =11),  # Style y-axis label
     axis.text.x = element_text(size = 12),  # Style x-axis ticks

### 2024-09-28 paper/paper.tex
added some comments, added sketches, organized

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Frequent Inspections Fail to Curb Violations in Toronto's Good-Standing Food Establishments},
  pdfauthor={Jerry Xia},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Frequent Inspections Fail to Curb Violations in Toronto's
Good-Standing Food Establishments\thanks{Code and data are available at:
\url{https://github.com/Jerryx2020/toronto_dinesafe_analysis}}}
\author{Jerry Xia}
\date{September 27, 2024}

\begin{document}
\maketitle
\begin{abstract}
This study analyzes inspection patterns and compliance outcomes using
Toronto's DineSafe dataset (2022-2024) to explore the relationship
between inspection frequency and food safety infractions in restaurants
and takeout establishments. The data reveals that restaurants have a
higher percentage of inspections resulting in infractions compared to
takeout establishments, highlighting more significant compliance
challenges in full-service operations. These findings suggest that
current regulatory practices may need to focus more on restaurants to
ensure public safety. By identifying gaps in inspection frequency and
infractions, this analysis highlights the need for more targeted
oversight to improve food safety.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\section{Introduction}\label{introduction}

In large urban centers like Toronto, ensuring food safety is a critical
public health priority. Foodborne illnesses pose significant risks, and
the health of the population is closely tied to the hygienic practices
of food establishments, including restaurants, food trucks, and takeout
locations. To address this, the DineSafe program, managed by Toronto
Public Health, enforces health and safety regulations by conducting
regular inspections of all food service establishments (Gelfand 2022).
These inspections result in outcomes ranging from a pass to a
conditional pass or even closure, depending on the establishment's
compliance with food safety standards. According to the Centers for
Disease Control and Prevention (CDC), frequent and transparent food
inspections are instrumental in reducing foodborne illness outbreaks, as
public posting of inspection results can encourage better compliance in
food service settings (Centers for Disease Control and Prevention 2024).
Despite this, significant variations exist in the frequency of
inspections and the severity of infractions, raising concerns about
whether some establishments, particularly mobile or temporary food
vendors, receive adequate regulatory attention (Analytics 2023).

This study focuses on analyzing Toronto's DineSafe dataset, covering
inspections from 2022 to 2024, to investigate patterns in inspection
frequency and compliance outcomes among different types of food
establishments. While previous research has broadly examined compliance
across food establishments, this paper delves deeper into the
relationships between establishment types---such as traditional
restaurants and takeout locations---and the severity of infractions
identified during inspections. Restaurants exhibit a higher rate of
severe infractions compared to takeout establishments, suggesting that
full-service operations may require more regulatory oversight to address
food safety challenges (Agency 2023).

To address this gap, the DineSafe inspection data for Toronto food
establishments was obtained and cleaned as described in
Section~\ref{sec-data-overview}. The results, presented in
Section~\ref{sec-data-results}, reveal key trends in the frequency of
inspections and the prevalence of infractions across different
establishment types. The findings indicate that while mobile food
vendors undergo fewer inspections than restaurants, they exhibit a
higher rate of severe infractions relative to the number of inspections
conducted. These insights, as discussed in Section~\ref{sec-discussion},
highlight the need for more frequent inspections or stricter regulations
in this sector to ensure public safety.

The remainder of this paper is organized as follows:
Section~\ref{sec-data-overview} provides a detailed overview of the data
and the methodology used to clean and analyze the dataset;
Section~\ref{sec-data-results} presents the key findings of the
analysis, including the inspection frequency and infraction severity;
and Section~\ref{sec-discussion} concludes with recommendations for
improving food safety oversight in Toronto. The appendix includes
additional materials, such as code and data, ensuring full
reproducibility of the results.

\section{Data}\label{sec-data}

\subsection{Overview}\label{sec-data-overview}

This analysis utilizes the DineSafe dataset from Toronto's Open Data
platform, accessed using the \texttt{opendatatoronto} package (Gelfand
2022). The dataset includes inspections conducted from 2022 to 2024,
providing detailed information on health inspections of food
establishments, including restaurants and takeout locations, throughout
Toronto. Key variables in the dataset include inspection dates,
infraction types, and establishment compliance statuses, which serve as
the primary indicators for assessing food safety. These regular
inspections are crucial for maintaining food hygiene standards, as
highlighted by the CDC, which emphasizes the importance of routine
inspections and the public posting of results, such as letter grades, to
encourage compliance and reduce foodborne illnesses (Centers for Disease
Control and Prevention 2024).

For this study, only the inspection data for ``Restaurant'' and ``Food
Take Out'' establishments that passed their most recent inspection were
retained. This focus allows for an in-depth exploration of the
correlation between inspection frequency and the occurrence of
infractions in establishments that are deemed to be in compliance. The
DineSafe dataset is updated regularly by Toronto Public Health, and is
considered open data under Toronto's Open Data Licence
(Section~\ref{sec-appendix-attribution}), as long as proper attribution
is provided (Toronto 2024).

The data analysis and processing were carried out using the R
programming language (R Core Team 2023), employing a range of
specialized packages. The \texttt{tidyverse} package (Hadley Wickham et
al. 2019) was used extensively for filtering, cleaning, and summarizing
the data. Additionally, the \texttt{ggplot2} package (H. Wickham 2016)
was applied to visualize patterns in inspection frequency and the
severity of infractions. Date-related data were managed using the
\texttt{lubridate} package (Grolemund and Wickham 2011) to ensure
consistency in handling inspection dates.

This analysis exclusively focuses on data for restaurants and takeout
establishments, as these are among the most common food service types in
Toronto, and represent a significant portion of the inspections carried
out by public health authorities. As described in
Section~\ref{sec-data-results}, the dataset was thoroughly cleaned and
prepared for analysis, ensuring that all irrelevant fields were removed
and the remaining data was formatted for accurate analysis.

\newpage

\subsection{Results}\label{sec-data-results}

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-inspection-vs-infraction-1.pdf}

}

\caption{\label{fig-inspection-vs-infraction}Inspection Count vs
Infraction Count per Establishment}

\end{figure}%

As shown in Figure~\ref{fig-inspection-vs-infraction}, the graph now
excludes only inspections where ``No Infraction'' was recorded. The
results show that establishments inspected more frequently tend to have
higher counts of infractions. This positive correlation reflects
findings from the literature that increased inspection frequency does
not necessarily reduce violations but instead highlights pre-existing
issues (Public Health 2023; Centers for Disease Control and Prevention
2024). The data was processed using the \texttt{tidyverse} package for
summarization and visualization (Hadley Wickham et al. 2019), while the
plots were generated using \texttt{ggplot2} (H. Wickham 2016).

\newpage

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-inspections-over-time-1.pdf}

}

\caption{\label{fig-inspections-over-time}Number of Inspections
Conducted Over Time (Monthly)}

\end{figure}%

As shown in Figure~\ref{fig-inspections-over-time}, the number of
inspections fluctuates month by month. Several external factors, such as
regulatory changes, seasonal variations, or public health crises like
the COVID-19 pandemic, likely influence these inconsistencies in
inspection activity (Analytics 2023). The visualization was created
using \texttt{ggplot2} (H. Wickham 2016) for plotting and
\texttt{lubridate} (Grolemund and Wickham 2011) for time-based grouping.

\newpage

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-restaurant-vs-takeout-1.pdf}

}

\caption{\label{fig-restaurant-vs-takeout}Percentage of Inspections
Resulting in Infractions: Restaurants vs.~Takeout}

\end{figure}%

As shown in Figure~\ref{fig-restaurant-vs-takeout}, the percentage of
inspections resulting in infractions is higher for restaurants (73.4\%)
compared to takeout establishments (58.6\%). This indicates that despite
takeout establishments being subject to fewer total inspections due to
their smaller numbers, restaurants have a greater percentage of
infractions per inspection. This could suggest that full-service
restaurants face more complex operational challenges that increase the
likelihood of non-compliance with food safety regulations. Research has
shown that full-service restaurants, due to the complexity of food
handling processes, tend to have higher rates of infractions compared to
smaller or more specialized establishments (Analytics 2023; Agency
2023). These findings underscore the need for continued regulatory focus
on restaurants, where maintaining compliance appears more difficult,
reflecting the operational differences between full-service restaurants
and takeout locations (Centers for Disease Control and Prevention 2024).

\newpage

\section{Discussion}\label{sec-discussion}

The analysis of the DineSafe data revealed several key insights into
food safety practices across Toronto's restaurants and takeout
establishments. One notable observation, as depicted in
Figure~\ref{fig-inspection-vs-infraction} and discussed in
Section~\ref{sec-data-results}, is the positive correlation between the
number of inspections conducted at an establishment and the number of
infractions identified. This finding suggests that more frequent
inspections tend to uncover more infractions. However, it raises the
question of whether these inspections are uncovering ongoing compliance
issues rather than fostering improvements. Similar conclusions have been
noted in previous studies, which found that increased inspection
frequency highlights existing compliance problems without necessarily
reducing the overall number of violations (Public Health 2023; Centers
for Disease Control and Prevention 2024).

When examining the number of inspections conducted over time, as shown
in Figure~\ref{fig-inspections-over-time}, and detailed in
Section~\ref{sec-data-results}, the data indicates inconsistencies in
inspection activity. These variations are likely influenced by external
factors such as regulatory changes, public health events (e.g., the
COVID-19 pandemic), or operational adjustments (Analytics 2023). While
spikes in inspection activity may correspond to heightened public health
concerns, ensuring sustained compliance requires consistent follow-up
actions. Without continuous enforcement measures, temporary increases in
inspections may not lead to long-term reductions in infractions.

Figure~\ref{fig-restaurant-vs-takeout} in Section~\ref{sec-data-results}
compares the percentage of inspections resulting in infractions between
restaurants and takeout establishments. Restaurants, with a 73.4\%
infraction rate compared to 58.6\% for takeout establishments
(Figure~\ref{fig-restaurant-vs-takeout}), face greater compliance
challenges, suggesting a need for enhanced regulatory scrutiny. This
suggests that full-service restaurants, due to their operational
complexity, may face greater compliance challenges with food safety
regulations. This observation is consistent with other research that
highlights the difficulties full-service restaurants experience in
maintaining hygiene and safety standards, as their more complex
operations involve multiple stages of food preparation and handling
(Analytics 2023; Agency 2023). These findings underscore the importance
of focused regulatory oversight on restaurants, where compliance
challenges appear more prevalent.

Despite the valuable insights provided by this analysis, certain
limitations must be acknowledged. As noted in
Section~\ref{sec-data-results}, the DineSafe dataset may not capture all
infractions, especially for establishments classified as low-risk and
inspected less frequently. This underrepresentation could result in an
incomplete view of food safety compliance across the city. Moreover,
some infractions may not lead to immediate enforcement actions (e.g.,
fines or closures), potentially allowing non-compliance to persist over
time. Addressing these limitations may require additional measures such
as follow-up inspections and broadening the range of establishments
subject to frequent scrutiny (Gelfand 2022).

Future research could investigate the application of machine learning
techniques to predict which establishments are most likely to fail
inspections based on historical data, allowing regulatory bodies to
allocate resources more efficiently (Public Health 2023). Additionally,
more detailed studies into the effectiveness of enforcement actions,
such as fines and mandatory re-inspections, could provide insights into
strategies for improving compliance rates. Comparative analyses between
cities with varying levels of transparency, such as public disclosure of
inspection scores, could further illuminate best practices for reducing
foodborne illness outbreaks and enhancing compliance, as discussed in
Section~\ref{sec-discussion}.

In conclusion, while the DineSafe program has been successful in
identifying non-compliant establishments, this study suggests that
frequent inspections alone are not enough to prevent food safety
violations. Targeted regulatory interventions, particularly in high-risk
and complex food service sectors like full-service restaurants, may be
required to strengthen compliance and safeguard public health in
Toronto.

\newpage

\appendix

\section{Appendix}\label{sec-appendix}

\subsection{Dataset and Graph Sketches}\label{sec-appendix-sketches}

Sketches depicting both the desired dataset and the graphs generated in
this analysis are available in the GitHub Repository.

\subsection{Data Cleaning}\label{sec-appendix-cleaning}

The data cleaning process was essential to prepare the raw DineSafe
dataset for accurate analysis. Initially, we filtered the data to focus
exclusively on ``Restaurant'' and ``Food Take Out'' establishments that
had passed their most recent inspection. This ensured that our analysis
targeted establishments in good standing, allowing us to assess how
inspection frequency correlates with violations in compliant
establishments.

Next, we removed irrelevant columns to simplify the dataset and enhance
clarity. For example, columns unrelated to inspection outcomes or
establishment types were excluded. Additionally, we ensured consistency
across the dataset by renaming certain columns for clarity and ease of
analysis.

To handle date-related data, we utilized the lubridate package
(Grolemund and Wickham 2011), which enabled consistent date formatting
and facilitated time-based analysis. The entire data cleaning process
was carried out using the tidyverse package (Hadley Wickham et al.
2019), which streamlined the filtering, mutating, and summarizing
operations essential for preparing the dataset for further exploration.

\subsection{Attribution Statement}\label{sec-appendix-attribution}

``Contains information licensed under the Open Government Licence --
Toronto'' (Toronto 2024).

\newpage

\section*{References}\label{sec-references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-fsa2023}
Agency, Food Standards. 2023. {``The Impact of Displaying Food Hygiene
Ratings on Compliance.''} \emph{FSA Research and Evidence}.
\url{https://science.food.gov.uk/}.

\bibitem[\citeproctext]{ref-hazel2023}
Analytics, Hazel. 2023. {``Post-Pandemic Food Safety Compliance: Key
Insights from Health Inspections.''} \emph{Nation's Restaurant News}.
\url{https://www.nrn.com/food-safety/post-pandemic-food-safety-compliance}.

\bibitem[\citeproctext]{ref-cdc2024}
Centers for Disease Control and Prevention. 2024. {``Posting Inspection
Scores Reduces Foodborne Illness Outbreaks in Restaurants.''} \emph{Food
Safety Insights}. \url{https://www.cdc.gov/foodsafety/}.

\bibitem[\citeproctext]{ref-citeOpendatatoronto}
Gelfand, D. 2022. \emph{Opendatatoronto: Access the Toronto Open Data
Portal}. \url{https://CRAN.R-project.org/package=opendatatoronto}.

\bibitem[\citeproctext]{ref-citeLubridate}
Grolemund, Garrett, and Hadley Wickham. 2011. \emph{Lubridate: Make
Dealing with Dates a Little Easier}.
\url{https://CRAN.R-project.org/package=lubridate}.

\bibitem[\citeproctext]{ref-mdpi2023}
Public Health, MDPI Journal of. 2023. {``Using Machine Learning to
Predict Non-Compliant Food Outlets.''} \emph{MDPI Open Access}.
\url{https://www.mdpi.com/journal/publichealth}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2023. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-tphlicense}
Toronto, City of. 2024. {``Open Government Licence -- Toronto.''}
\url{https://open.toronto.ca/open-data-license/}.

\bibitem[\citeproctext]{ref-citeGgplot2}
Wickham, H. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}.
Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\bibitem[\citeproctext]{ref-citeTidyverse}
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\end{CSLReferences}

\end{document}

### 2024-09-28 scripts/00-simulate_data.R

@@ -8,6 +8,7 @@
 # Any other information needed? None.
 
 #### Workspace setup ####
+# install.packages("tidyverse")
 library(tidyverse)
 
 #### Simulate data ####

### 2024-09-28 scripts/01-download_data.R

@@ -9,6 +9,8 @@
 
 #### Workspace setup ####
 # Load required libraries
+# install.packages("opendatatoronto")
+# install.packages("tidyverse")
 library(opendatatoronto)
 library(tidyverse)        
 

### 2024-09-28 scripts/02-data_cleaning.R

@@ -8,6 +8,7 @@
 # Any other information needed? None.
 
 #### Workspace setup ####
+# install.packages("tidyverse")
 library(tidyverse)
 
 # Load raw data and exclude the problematic "Min. Inspections Per Year" column (8th column): contains char where expected is a double

### 2024-09-28 scripts/03-test_data.R

@@ -8,6 +8,7 @@
 # Any other information needed? None.
 
 #### Workspace setup ####
+# install.packages("tidyverse")
 library(tidyverse)
 
 #### Test data ####

### 2024-09-28 paper/paper.qmd
finialized paper

@@
 ---
-title: "Frequent Inspections Fail to Curb Violations in Toronto's Good-Standing Food Establishments"
+title: "Inspection Frequency Reveals Violation Patterns in Toronto Restaurants and Takeout Establishments (2022-2024)"
 author: 
@@
 thanks: "Code and data are available at: <https://github.com/Jerryx2020/toronto_dinesafe_analysis>"
-date: "`r format(Sys.Date() - 1, '%A, %B %d, %Y')`"
+date: "`r format(Sys.Date() - 1, '%A, %B %d, %Y')`" #chatgpt assisted how to create a custom date for due date Sept 27
 date-format: long
-abstract: "This study analyzes inspection patterns and compliance outcomes using Toronto’s DineSafe dataset (2022-2024) to explore the relationship between inspection frequency and food safety infractions in restaurants and takeout establishments.
-The data reveals that restaurants have a higher percentage of inspections resulting in infractions compared to takeout establishments, highlighting more significant compliance challenges in full-service operations.
-These findings suggest that current regulatory practices may need to focus more on restaurants to ensure public safety.
-By identifying gaps in inspection frequency and infractions, this analysis highlights the need for more targeted oversight to improve food safety.
-"
+abstract: "This study analyzes Toronto’s DineSafe dataset (2022-2024) to examine the relationship between inspection frequency and food safety infractions in restaurants and takeout establishments.
+Results show 73.4% of restaurant inspections lead to infractions, compared to 58.6% for takeout locations, with more frequent inspections correlating with higher infraction counts.
+These findings suggest that regulatory efforts may need to be more targeted, particularly in full-service operations, to address the higher violation rates and improve overall food safety compliance in Toronto."
 format: pdf
 number-sections: true
-toc: true #https://quarto.org/docs/reference/formats/pdf.html
+toc: true #chatgpt asissted with table of contents
 bibliography: references.bib
@@
 
-In large urban centers like Toronto, ensuring food safety is a critical public health priority.
-Foodborne illnesses pose significant risks, and the health of the population is closely tied to the hygienic practices of food establishments, including restaurants, food trucks, and takeout locations.
-To address this, the DineSafe program, managed by Toronto Public Health, enforces health and safety regulations by conducting regular inspections of all food service establishments [@citeOpendatatoronto].
-These inspections result in outcomes ranging from a pass to a conditional pass or even closure, depending on the establishment's compliance with food safety standards.
-According to the Centers for Disease Control and Prevention (CDC), frequent and transparent food inspections are instrumental in reducing foodborne illness outbreaks, as public posting of inspection results can encourage better compliance in food service settings [@cdc2024].
-Despite this, significant variations exist in the frequency of inspections and the severity of infractions, raising concerns about whether some establishments, particularly mobile or temporary food vendors, receive adequate regulatory attention [@hazel2023].
+In large urban centers like Toronto, food safety is a critical public health concern, as foodborne illnesses can pose significant risks to the population.
+The DineSafe program, managed by Toronto Public Health, enforces health and safety regulations by conducting regular inspections of food establishments, including restaurants, food trucks, and takeout locations.
+These inspections result in outcomes such as a pass, conditional pass, or closure, depending on compliance with food safety standards [@citeOpendatatoronto].
+According to the Centers for Disease Control and Prevention (CDC), regular and transparent food inspections are essential for reducing foodborne illness outbreaks, as public posting of results encourages compliance [@cdc2024].
+However, disparities in inspection frequency and the severity of infractions across establishment types, particularly between mobile vendors and traditional restaurants, raise concerns about whether current regulatory efforts are sufficient [@hazel2023].
 
-This study focuses on analyzing Toronto’s DineSafe dataset, covering inspections from 2022 to 2024, to investigate patterns in inspection frequency and compliance outcomes among different types of food establishments.
-While previous research has broadly examined compliance across food establishments, this paper delves deeper into the relationships between establishment types—such as traditional restaurants and takeout locations—and the severity of infractions identified during inspections.
-Restaurants exhibit a higher rate of severe infractions compared to takeout establishments, suggesting that full-service operations may require more regulatory oversight to address food safety challenges [@fsa2023].
+This paper analyzes the DineSafe dataset from 2022 to 2024 to explore patterns in inspection frequency and compliance outcomes across different types of food establishments.
+While previous studies have examined broad compliance trends, this research focuses specifically on how the type of establishment—such as restaurants versus takeout locations—affects the likelihood of infractions.
+Notably, restaurants tend to have more infractions compared to takeout establishments, suggesting that more stringent oversight may be needed for full-service operations [@fsa2023].
+Additionally, restaurant vendors exhibit high rates of infractions relative to the number of inspections conducted, further underscoring potential regulatory gaps.
 
-To address this gap, the DineSafe inspection data for Toronto food establishments was obtained and cleaned as described in @sec-data-overview.
-The results, presented in @sec-data-results, reveal key trends in the frequency of inspections and the prevalence of infractions across different establishment types.
-The findings indicate that while mobile food vendors undergo fewer inspections than restaurants, they exhibit a higher rate of severe infractions relative to the number of inspections conducted.
-These insights, as discussed in @sec-discussion, highlight the need for more frequent inspections or stricter regulations in this sector to ensure public safety.
+To address the variations in inspection frequency and compliance, the DineSafe inspection data for Toronto's food establishments was cleaned and analyzed, as described in @sec-data-overview and @sec-appendix-cleaning.
+The results, detailed in @sec-data-results, show that establishments inspected more frequently tend to have higher infraction counts, suggesting that frequent inspections uncover more violations but may not effectively reduce them (@fig-inspection-vs-infraction).
+Inspection activity fluctuated monthly from 2022 to 2024, with peaks likely driven by public health crises or regulatory changes (@fig-inspections-over-time).
+Restaurants, in particular, exhibit a higher percentage of inspections resulting in infractions (73.4%) compared to takeout establishments (58.6%), highlighting the greater compliance challenges faced by full-service operations (@fig-restaurant-vs-takeout).
+These findings emphasize the need for targeted regulatory efforts to improve food safety compliance across different sectors.
 
-The remainder of this paper is organized as follows: @sec-data-overview provides a detailed overview of the data and the methodology used to clean and analyze the dataset; @sec-data-results presents the key findings of the analysis, including the inspection frequency and infraction severity; and @sec-discussion concludes with recommendations for improving food safety oversight in Toronto.
-The appendix includes additional materials, such as code and data, ensuring full reproducibility of the results.
+The remainder of this paper is organized as follows: @sec-data-overview provides a detailed overview of the dataset and methodology; @sec-data-results presents the key findings, including inspection frequency and @sec-discussion concludes with recommendations for improving food safety oversight in Toronto.
+An appendix with supplementary materials, including cleaning and sketches, ensures full reproducibility of the results.
 
@@
 The dataset includes inspections conducted from 2022 to 2024, providing detailed information on health inspections of food establishments, including restaurants and takeout locations, throughout Toronto.
- Key variables in the dataset include inspection dates, infraction types, and establishment compliance statuses, which serve as the primary indicators for assessing food safety.
+
+The DineSafe dataset used in this study contains 20,481 individual inspections from 2022 to 2024, covering 6,231 unique food establishments across Toronto.
+Among these establishments, 73.5% are categorized as restaurants, while the remaining 26.5% are takeout establishments.
+The average duration between consecutive inspections for any establishment is 170 days, highlighting a relatively consistent inspection schedule.Of the infractions recorded, 37.2% are categorized as "Significant", while 17.5% are "Crucial", indicating a lower proportion of severe food safety violations.
+Only 45.3% of recorded infractions are labeled "Minor".
+
+The key variables in the dataset include:
+
+- **Establishment ID**: A unique identifier for each food establishment in Toronto.
+- **Establishment Type**: The type of food establishment, such as "Restaurant" or "Food Take Out." This variable helps categorize the establishments and compare compliance across different types of food services.
+- **Severity**: This variable indicates the severity of food safety infractions, categorized as "C - Critical," "S - Significant," "M - Minor," and "No Infraction." Critical infractions represent immediate risks to public health, while minor infractions are less severe.
+- **Inspection Date**: The date on which an inspection occurred. This variable allows for temporal analysis of inspections over time, such as monthly or yearly trends.
+- **Establishment Status**: Indicates whether the establishment passed, failed, or received a conditional pass following an inspection.
+
+They serve as the primary indicators for assessing food safety.
 These regular inspections are crucial for maintaining food hygiene standards, as highlighted by the CDC, which emphasizes the importance of routine inspections and the public posting of results, such as letter grades, to encourage compliance and reduce foodborne illnesses [@cdc2024].
 
+Several variables required construction or manipulation to facilitate the analysis.
+Counting Infractions: For the analysis of infraction rates, the Severity variable was used to construct a count of confirmed infractions for each establishment.
+Specifically, any row where the Severity was not labeled as "No Infraction" was considered to have a confirmed infraction.
+This constructed variable allowed for a clearer view of the total number of infractions per establishment.
+Infraction Percentage: To compare infraction rates between establishment types (restaurants vs.
+takeout), we calculated the Percentage of Inspections with Infractions.
+This involved dividing the total number of inspections where an infraction occurred by the total number of inspections for each establishment type.
+This metric allowed us to assess compliance rates across different food service sectors.
+
+DineSafe is subject to limitations in measurement.
+For example, while most severe infractions are likely to be well-documented, less serious violations may be underreported, or follow-up inspections may not be conducted in a timely manner.
+This is a limitation noted in broader studies of food safety compliance systems [@hazel2023].
+Other relevant datasets in Toronto, such as the Toronto Public Health Inspection Results and COVID-19 Business and Community Closure Data, were considered but not used due to their broader scope or focus on closures rather than detailed food safety inspections.
+While these datasets offer insights into public health compliance, they lack the detailed inspection frequency and infraction data that the DineSafe dataset provides [@citeOpendatatoronto; @tphlicense].
+The DineSafe dataset offers detailed insights into food safety but has limitations.
+It may underrepresent infractions in lower-risk establishments, which are inspected less frequently, as previous studies indicate that inspections tend to focus on higher-risk venues [@hazel2023].
+Furthermore, not all infractions lead to immediate penalties, meaning some compliance issues may persist undetected [@hazel2023; @fsa2023].
+The dataset also does not capture asymptomatic outbreaks or unreported violations, which can affect public health [@fsa2023].
+
 For this study, only the inspection data for "Restaurant" and "Food Take Out" establishments that passed their most recent inspection were retained.
+This analysis exclusively focuses on data for restaurants and takeout establishments, as these are among the most common food service types in Toronto.
 This focus allows for an in-depth exploration of the correlation between inspection frequency and the occurrence of infractions in establishments that are deemed to be in compliance.
 The DineSafe dataset is updated regularly by Toronto Public Health, and is considered open data under Toronto’s Open Data Licence (@sec-appendix-attribution), as long as proper attribution is provided [@tphlicense].
-
-The data analysis and processing were carried out using the R programming language [@citeR], employing a range of specialized packages.
-The `tidyverse` package [@citeTidyverse] was used extensively for filtering, cleaning, and summarizing the data.
-Additionally, the `ggplot2` package [@citeGgplot2] was applied to visualize patterns in inspection frequency and the severity of infractions.
-Date-related data were managed using the `lubridate` package [@citeLubridate] to ensure consistency in handling inspection dates.
-
-This analysis exclusively focuses on data for restaurants and takeout establishments, as these are among the most common food service types in Toronto, and represent a significant portion of the inspections carried out by public health authorities.
-As described in @sec-data-results, the dataset was thoroughly cleaned and prepared for analysis, ensuring that all irrelevant fields were removed and the remaining data was formatted for accurate analysis.
-
 \newpage
@@
 ## Results {#sec-data-results}
-
 ```{r}
@@
   geom_point(color = "purple", size = 2, alpha = 0.7) +  # Purple dots
-  geom_smooth(method = "lm", color = "darkgray", linetype = "dashed", se = FALSE) +  # Regression line
+  geom_smooth(method = "lm", color = "darkgray", linetype = "dashed", se = FALSE) +  # chatgpt regression line
   scale_x_continuous(breaks = seq(0, 60, by = 10)) +  
@@
 This positive correlation reflects findings from the literature that increased inspection frequency does not necessarily reduce violations but instead highlights pre-existing issues [@mdpi2023; @cdc2024].
-The data was processed using the `tidyverse` package for summarization and visualization [@citeTidyverse], while the plots were generated using `ggplot2` [@citeGgplot2].
 
@@
   geom_line(color = "black", size = 1.2) +  # Grayscale line for monthly inspections
-  geom_point(color = "purple", size = 2, alpha = 0.8) +  # Purple points for each month
+  geom_point(color = "purple", size = 2, alpha = 0.8) + 
   labs(x = "Month",
@@
     axis.text.y = element_text(size = 10, color = "black"),  # Style y-axis ticks
-    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # Lighter grid lines
+    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # grid lines
     panel.grid.minor = element_blank()  # Remove minor grid lines
@@
 Several external factors, such as regulatory changes, seasonal variations, or public health crises like the COVID-19 pandemic, likely influence these inconsistencies in inspection activity [@hazel2023].
-The visualization was created using `ggplot2` [@citeGgplot2] for plotting and `lubridate` [@citeLubridate] for time-based grouping.
 
@@
 ggplot(type_percentage_comparison, aes(x = `Establishment Type`, y = percentage_with_infractions)) +
-  geom_bar(stat = "identity", fill = "gray40", color = "black", width = 0.6) +  # Grayscale bar with refined styling
+  geom_bar(stat = "identity", fill = "gray40", color = "black", width = 0.6) +  # Grayscale bar
   geom_text(aes(label = sprintf("%.1f%%", percentage_with_infractions)), vjust = -0.5, size = 5, color = "purple") +  # Purple labels
@@
     axis.text.y = element_text(size = 12),  # Style y-axis ticks
-    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # Lighter grid lines
+    panel.grid.major = element_line(size = 0.2, color = "gray80"),  # grid lines
     panel.grid.minor = element_blank()  # Remove minor grid lines
   )
+
+#chatgpt learned how to group by type
 
@@
 These findings underscore the need for continued regulatory focus on restaurants, where maintaining compliance appears more difficult, reflecting the operational differences between full-service restaurants and takeout locations [@cdc2024].
+All plots were generated using `ggplot2` [@citeGgplot2].
 
@@
 Despite the valuable insights provided by this analysis, certain limitations must be acknowledged.
-As noted in @sec-data-results, the DineSafe dataset may not capture all infractions, especially for establishments classified as low-risk and inspected less frequently.
-This underrepresentation could result in an incomplete view of food safety compliance across the city.
-Moreover, some infractions may not lead to immediate enforcement actions (e.g., fines or closures), potentially allowing non-compliance to persist over time.
-Addressing these limitations may require additional measures such as follow-up inspections and broadening the range of establishments subject to frequent scrutiny [@citeOpendatatoronto].
+The DineSafe dataset captures only reported infractions during scheduled inspections, and some minor infractions may go unreported due to limited follow-up or enforcement in lower-risk establishments.
+For instance, Hazel Analytics noted in their 2023 report that establishments classified as low-risk may experience fewer inspections, potentially underrepresenting compliance issues in these sectors [@hazel2023].
+This can be easily seen in the data where quick sorting by specific columns results in many thousands of blank entries.
+Additionally, infractions that do not lead to immediate enforcement actions, such as fines or temporary closures, may allow non-compliance to persist undetected for extended periods [@cdc2024].
+Addressing these limitations will likely require more frequent and unannounced follow-up inspections, as well as extending scrutiny to a wider range of establishments [@fsa2023].
 
-Future research could investigate the application of machine learning techniques to predict which establishments are most likely to fail inspections based on historical data, allowing regulatory bodies to allocate resources more efficiently [@mdpi2023].
-Additionally, more detailed studies into the effectiveness of enforcement actions, such as fines and mandatory re-inspections, could provide insights into strategies for improving compliance rates.
-Comparative analyses between cities with varying levels of transparency, such as public disclosure of inspection scores, could further illuminate best practices for reducing foodborne illness outbreaks and enhancing compliance, as discussed in @sec-discussion.
+Future research could investigate the application of machine learning techniques to predict establishments most at risk of failing inspections based on historical data, enabling more efficient allocation of resources [@mdpi2023].
+Further studies examining the impact of enforcement actions, such as fines and mandatory re-inspections, on long-term compliance would provide valuable insights into regulatory effectiveness.
+Comparative analyses between cities with varying levels of transparency—such as those that publicly disclose inspection scores—could offer additional insights into effective strategies for reducing foodborne illness outbreaks and improving compliance [@fsa2023].
 
-In conclusion, while the DineSafe program has been successful in identifying non-compliant establishments, this study suggests that frequent inspections alone are not enough to prevent food safety violations.
-Targeted regulatory interventions, particularly in high-risk and complex food service sectors like full-service restaurants, may be required to strengthen compliance and safeguard public health in Toronto.
+In conclusion, while the DineSafe program has been effective in identifying non-compliant establishments, this study suggests that frequent inspections alone are insufficient to prevent recurring food safety violations.
+Inconsistent inspection schedules, particularly among high-risk establishments, may further undermine the efficacy of regulatory efforts.
+To safeguard public health, regulatory bodies should consider implementing more targeted interventions, such as increased oversight of full-service restaurants and sectors with high infraction rates, coupled with stricter penalties for non-compliance [@hazel2023; @fsa2023].
 

### 2024-09-28 scripts/03-test_data.R
test

@@ -24,7 +24,7 @@ missing_values <- data %>%
     missing_inspection_date = sum(is.na(`Inspection Date`))
   )
 
-print(missing_values)  # Check if there are missing values in critical columns
+print(missing_values)  # check if there are missing values in critical columns
 
 # 2. Test for invalid severity values
 # Include "No Infraction" as a valid severity level, as well as other valid levels.
@@ -39,7 +39,7 @@ print(invalid_severity_values)  # Any rows where Severity is not a valid categor
 complete_cases_check <- data %>%
   filter(is.na(`Establishment ID`) | is.na(`Establishment Type`) | is.na(Severity) | is.na(`Inspection Date`))
 
-print(complete_cases_check)  # Output rows where any of the key fields are missing
+print(complete_cases_check)  # output rows where any of the key fields are missing
 
 # Test Results Summary
 print("Sanity checks completed for DineSafe data.")
